# Create simple vann diagram

```{r}
#| warning: false
#| message: false
pacman::p_load(ggVennDiagram,
               ggplot2)
```

## Basic 2D Vann diagram:

```{r}
#| warning: false
#| message: false
x <- list(A = 1:5, B = 2:7)
ggVennDiagram(x)
```

## 3D Venn diagram:

If you pass a list with three elements you will get some Venn diagram like the following

```{r}
#| warning: false
#| message: false
x <- list(A = 1:5, B = 2:7, C = 5:10)
ggVennDiagram(x)
```

## Changing the colors of the diagram

-   As `ggVennDiagram` is based on `ggplot2` you can add more layers or override the ones existing. In order to change the colors use `scale_fill_gradient` as in the following example:

```{r}
x <- list(A = 1:5, B = 2:7, C = 5:10)
ggVennDiagram(x) + 
  scale_fill_gradient(low = "#F4FAFE", high = "#4981BF") 
```

-   The border can also be customized but making use of color argument. The line width and line style can be changed with `lwd` and `lty`, respectively.

## Labels and group names

```{r}
x <- list(A = 1:5, B = 2:7, C = 5:10)
ggVennDiagram(x, category.names = c("A",
                                    "B",
                                    "C"))
```

-   Labels with percentages

```{r}
x <- list(A = 1:5, B = 2:7, C = 5:10)
ggVennDiagram(x, 
              label = "percent",
              category.names = c("A",
                                    "B",
                                    "C")) 
```

-   Remove the labels

```{r}
x <- list(A = 1:5, B = 2:7, C = 5:10)
ggVennDiagram(x, 
              label = "none",
              category.names = c("A",
                                    "B",
                                    "C")) 
```

-Customize the legend

```{r}
x <- list(A = 1:5, B = 2:7, C = 5:10)
ggVennDiagram(x) + 
  guides(fill = guide_legend(title = "Title")) +
  theme(legend.title = element_text(color = "black"),
        legend.position = "bottom") 
```

## 4D Venn diagram

You can also pass a list with up to four different elements.

```{r}
x <- list(A = 1:5, B = 2:7, C = 5:10, D = 8:15)
ggVennDiagram(x)
```
