# Difference in Differences

```{r}
pacman::p_load(tidyverse,
               knitr,
               dplyr,
               tidyr,
               ggplot2,
               readr,
               kableExtra,
               stargazer,
               lmtest,
               plm)
```

## Wide and long data formats

Wide data set:

```{r}
city_wide_df <- read.csv("data/city_wide.csv")
city_wide_df %>% knitr::kable() %>% kableExtra::kable_styling()
```

```{r}
city_long_df <- read.csv("data/city_long.csv")
city_long_df %>% knitr::kable() %>% kableExtra::kable_styling()
```

Convert "city_wide_df" to long data set

```{r}
city_wide_df %>%
  tidyr::pivot_longer(
    cols = c(pop_2000, pop_2010, pop_2020), # -city, !city, starts_with(pop_), etc... would also work
    names_to = "year", # where do we want the names of the columns to go? (year)
    names_prefix = "pop_", # names_prefix removes matching text from the start of each variable name (not always necessary)
    values_to = "pop" # where do we want the values in the columns to go? (pop)
  )
```

### Why do we care about the data format

In some instances long format datasets are required for advanced statistical analysis and graphing. For example, if we wanted to run the regression formulation of the difference in differences model, we would need to input our data in long format. Furthermore, having our data in long format is very useful when plotting. Packages such as ggplot2, expect that your data will be in long form for the most part.

```{r}
soda_tax_df <- read.csv("data/soda_tax_df.csv")
```


Are this wide or long data

```{r}
soda_tax_df %>% head(10)
```

Our soda_tax_df is in wide format. We can convert our data to a long format to render the time and treatment dummy variables and save is to the soda_tax_df_long.

We will utilize the pivot_longer() function from tidyr to format our data frame.

```{r}
soda_tax_df_long <- 
  soda_tax_df %>% # the wide format df
  tidyr::pivot_longer(cols = c(pre_tax, post_tax), # both contain information about soda drank at two points in time
                      names_to = "period", # grab the names of pre and post and save them to period
                      values_to = "soda_drank") %>% # grab values from pre and post and put them in soda_drank
  dplyr::mutate(after_tax = ifelse(period == "post_tax", 1, 0)) # create dummy for period

head(soda_tax_df_long, 10)
```

### Exploring our data

We can use our soda_tax_df to explore the distribution of soda consumption at different points in time.

Let's try first to look at the differences in the distribution only at the

#### pre-tax time period:

```{r}
ggplot(soda_tax_df, aes(x = pre_tax, fill = factor(treatment))) + 
  geom_density(alpha = 0.5) + # density plot with transparency (alpha = 0.5)
  scale_fill_manual(name = " ", # changes to fill dimension
                     values = c("#a7a8aa", "#cc0055"),
                     labels = c("Control", "Treatment")) +
  theme_minimal() +
  theme(legend.position = "bottom") +
  labs(title = "Distribution of soda consumption before the tax was imposed",
       x = "Soda consumtion (oz)",
       y = "Density")
```

Let's look at the

#### post-tax period:

```{r}
ggplot(soda_tax_df, aes(x = post_tax, fill = factor(treatment))) + 
  geom_density(alpha = 0.5) + # density plot with transparency (alpha = 0.5)
  scale_fill_manual(name = " ", # changes to fill dimension
                     values = c("#a7a8aa", "#cc0055"),
                     labels = c("Control", "Treatment")) +
  theme_minimal() +
  theme(legend.position = "bottom") +
  labs(title = "Distribution of soda consumption after the tax was imposed",
       x = "Soda consumtion (oz)",
       y = "Density")
```

Since in our soda_tax_df_long we represent the time and soda consumption dimensions under the same columns, we can create even more complex graphs.

#### grammar of graphs: Facets

We will use facet_grid() which forms a matrix of panels defined by row and column faceting variables. It is most useful when you have two discrete variables, and all combinations of the variables exist in the data.

```{r}
soda_tax_df_long %>% 
  dplyr::mutate(period = ifelse(period == "post_tax", "T1 - Post-tax", "T0 - Pre-tax"), # create more meaningful labels
                treatment = ifelse(treatment == 1, "Treated (D=1)", "Untreated (D=0)")) %>%
  dplyr::group_by(period, treatment) %>% # group to extract means of each group at each time
  dplyr::mutate(group_mean = mean(soda_drank)) %>% # extract means of each group at each time
ggplot(., aes(x = soda_drank, fill = factor(treatment))) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(name = " ", # changes to fill dimension
                     values = c("#cc0055", "#a7a8aa"),
                     labels = c("Treatment", "Control")) +
  facet_grid(treatment~period) + # we specify the matrix (treatment and period)
  geom_vline(aes(xintercept = group_mean), linetype = "longdash") + # add vertical line with the mean
  theme_bw() +
  theme(legend.position = "none") +
  labs(x = "Soda consumed (oz)",
       y = "Density")
```

### Modeling and estimating

So far we have ignored time in our estimations. Difference-in-differences compares the changes in outcomes over time between units under different treatment states. This allows us to correct for any differences between the treatment and comparison groups that are constant over time assuming that the trends in time are parallel.

#### Calculating without time

If we did not have the pre_tax baseline measure, we would likely utilize the post_tax to explore the average effect on the treated. In this case, we would model this as:

```{r}
after_model <- lm(post_tax ~ treatment, data = soda_tax_df)
stargazer(after_model, type = "text")
```

We could read this result substantively as: those who lived in districts were the tax was implemented consumed on average 146.9 ounces less of sugar-added drinks per week compared to those who lived in districts were the tax was not put in place. This calculation would give us a comparison of the treatment and control groups after treatment.

We would have to think carefully about possible factors that could differentiate our treatment and control groups. We use a treatment indicator based on the districts where the measure was able to be implemented. Treatment was not fully randomly assigned, so there may be lots of potential confounders that create baseline differences in the scores for those living in Old Eagleton compared to those in Snake Lounge, which also affect the after-treatment comparisons.

If we think about the mechanics behind this naive calculation, we are just comparing the average observed outcomes for those treated and not treated after the tax was imposed:

| Treatment | Average after tax |
|-----------|-------------------|
| 0         | 523.2726          |
| 1         | 376.3548          |

```{r}
ggplot(soda_tax_df, aes(x = post_tax, fill = factor(treatment))) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(name = " ", # changes to fill dimension
                     values = c("#a7a8aa", "#cc0055"),
                     labels = c("Control", "Treatment")) +
  geom_vline(xintercept = 523.27, linetype = "longdash", color = "#a7a8aa") + #avg for the untreated
  geom_vline(xintercept = 376.35, linetype = "longdash", color = "#cc0055") + #avg for the treated
  theme_minimal() +
  theme(legend.position = "bottom") +
  labs(title = "Distribution of soda consumption after the tax was imposed",
       x = "Soda consumtion (oz)",
       y = "Density")
```

#### Including the time dimension (Manual extraction of the DiD estimate)

| Treatment | Pre-tax | Post-tax | Difference |
|-----------|---------|----------|------------|
| 1         | 511.13  | 376.35   | -134.78    |
| 0         | 508.31  | 523.27   | 14.97      |

We can just manually substract:

$$ DiD = -134.79 -14.97 = -149.76$$

#### Including the time dimension

We can introduce the time component to our calculation by incorporating the pre-treatment levels of sugar-added drink consumption, which gives us the diff-in-diff estimand. We could calculate this in a fairly straightforward manner by creating a variable assessing the change in our wide format data frame:

-   change: The difference in sugar-added drink consuption between post- and pre-tax

```{r}
#| warning: false
soda_tax_df <- soda_tax_df %>%
  dplyr::mutate(change = post_tax - pre_tax) #simple substraction

did_model <- lm(change ~ treatment, data = soda_tax_df)
stargazer(did_model, after_model, type = "text")
```

We could read this result substantively as: those who lived in districts were the tax was implemented consumed on average 149.7 ounces less of sugar-added drinks per week compared to those who lived in districts were the tax was not put in place. This calculation would give us the change, or difference, in sugar-added drink consumption for treatment and control groups.

To believe the results of our did_model, we would need to believe that there are parallel trends between the two groups.

#### Including the time dimension (Regression formulation of the DiD model)

We can re-format our data to gather our diff-in-diff estimate

$$ Y_{it} = ......$$

For this calculation we need our data in long format to use the time and treatment dummy variables. We can see that under our long format, we have two entries for every individual.

```{r}
#| warning: false
did_long <- lm(soda_drank ~ treatment + after_tax + treatment*after_tax, data = soda_tax_df_long) #running our model

did_long_clustered_se <- coeftest(did_long, vcov=vcovHC(did_long,type="HC0",cluster="district")) #clustering out standard errors at the district level

stargazer::stargazer(did_long_clustered_se, type = "text")
```

```{r}
soda_tax_df_long %>% 
  dplyr::mutate(period = ifelse(period == "post_tax", "T1 - Post-tax", "T0 - Pre-tax"), # create more meaningful labels
                treatment = ifelse(treatment == 1, "Treated (D=1)", "Untreated (D=0)")) %>%
  dplyr::group_by(period, treatment) %>% # group to extract means of each group at each time
  dplyr::mutate(group_mean = mean(soda_drank)) %>% # extract means of each group at each time
ggplot(., aes(x = soda_drank, fill = factor(treatment))) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(name = " ", # changes to fill dimension
                     values = c("#cc0055", "#a7a8aa"),
                     labels = c("Treatment", "Control")) +
  facet_grid(treatment~period) + # we specify the matrix (treatment and period)
  geom_vline(aes(xintercept = group_mean), linetype = "longdash") + # add vertical line with the mean
  theme_bw() +
  theme(legend.position = "none") +
  labs(x = "Soda consumed (oz)",
       y = "Density")
```

```{r}
soda_tax_df_long %>%
  dplyr::group_by(period, treatment) %>% # group to extract means of each group at each time
  dplyr::mutate(group_mean = mean(soda_drank)) %>%
  ggplot(aes(x = after_tax, y = group_mean, color = factor(treatment))) +
  geom_point() +
  geom_line(aes(x = after_tax, y = group_mean)) +
  scale_x_continuous(breaks = c(0,1)) +
  scale_color_manual(name = " ", # changes to color dimension
                     values = c("#a7a8aa", "#cc0055"),
                     labels = c("Control", "Treatment")) +
  labs(x = "Time periods", y = "Ounces of soda drank per week", color = "Treatment group")+
  theme_minimal() 
```

-   Drafting some brief recommedations

Based on your analysis of the data at hand, you decide to recommend that the tax measure should move forward in the rest of Pawnee. You state that it is a very good example of a pigouvian tax, which captures the negative externalities not included in the market price of sugar-added drinks. The findings suggest that the tax reduced the weekly sugar-added drink consumption by about 150 luquid ounces (almost 4.5 liters).