# Visualization of qualitative data

Visualization can either be made with vectors and tables but also though diagrams like histograms, bar charts, pie charts etc. In each case, it is quite important to distinct our observations in quantitative or qualitative data. We start with qualitative data, which are usually represented with tables, bar charts and pie charts.

Example: Let’s assume we have the answers given by 20 persons to the question: “What means of transportation do you use every day to go to your job?”. Respondents had to choose between car, bus, metro and foot. The answers given where:

> car, car, bus, metro, metro, car, metro, metro, foot, car, foot, bus, bus, metro, metro, car, car, car, metro, car

We initially insert data by using the m vector. Our data are categorical so we should add these values inside double quotes:

```{r}
m = c("car", "car", "bus", "metro", "metro", "car", "metro", "metro", "foot", "car", "foot", "bus", "bus", "metro", "metro", "car", "car", "car", "metro", "car")
```

FREQUENCY TABLE: We can represent our data in a frequency table by using the table() command.

```{r}
table(m)
```

If we wanted to view the relative frequency then we could use the prop.table(table()) command.

```{r}
prop.table(table(m))
```

BAR CHARTS: Frequency bar charts can be displayed with the command:

```{r}
barplot(table(m))
```

Relative frequency bar charts can be displayed with the command:

```{r}
barplot(prop.table(table(m)))
```

PIE CHART: We can display our data in a pie chart by using the command:

```{r}
pie(table(m))
```

### Contigency Matrix

A contingency matrix has to do with two categorical variables and displays their frequency distribution.

-   Let’s assume that in the previous example’s data we have added the gender of each person.

-   For convenience, let’s assume that the first 8 people were male (M) and the remaining 12 were female (F).

We create the vector g:

```{r}
g = c(rep("M", 8), rep("F", 12))
g
```

We create the double input frequency table mg like this:

```{r}
mg = table(m,g)
mg
```

Then, using the table, we calculate marginal frequencies in terms of means of transportation and gender, respectively:

```{r}
margin.table(mg,1)
```

```{r}
margin.table(mg,2)
```

We can work with the relative frequency table with the exact same way.

```{r}
prop.table(mg)
```

```{r}
prop.table(mg,1)
```

```{r}
prop.table(mg,2)
```

### Stacked bar charts and grouped bar charts

We can display qualitative data coming from the values of two variables by using a stacked bar chart or a grouped bar chart.

```{r}
barplot(mg)
```

By using the command parameters, the result is much better:

```{r}
barplot(mg, xlim=c(0,2), xlab="Gender",
        legend =levels(m), col =1:4)
```

```{r}
barplot(mg, xlim = c(0,2), xlab = "Transportation",
        legend = levels(g), col=1:2)
```

The below commands create grouped bar charts.

```{r}
barplot(prop.table(mg,1), width = 0.25,
        xlim = c(0,3), ylim = c(0,1), xlab = "Gender",
        legend = levels(m), beside = T, col = 1:4)
```

```{r}
mg = table(g,m)
barplot(prop.table(mg,2), width = 0.25,
       xlim = c(0,3), ylim = c(0,1), xlab = "Transportation",
       legend = levels(g), beside = T, col = 1:2)
```
